Spike spectra for recurrences

### 2022-11-04 Manuscript/inter_spike_manuscript_mdpi.tex
Very last polishing before resubmission

% Abstract (Do not insert blank lines, i.e. \\) 
\abstract{In recurrence analysis, the $\tau$-recurrence rate encodes the periods of the cycles of the underlying high-dimensional time series. It thus plays a similar role to the autocorrelation for scalar time-series in encoding temporal correlations. However, its Fourier decomposition does not have a clean interpretation. Thus there is no satisfactory analogue to the power spectrum in recurrence analysis. In this paper we introduce a novel way to decompose the $\tau$-recurrence rate using an over-complete basis of Dirac combs together with a sparsity regularization. We show that this decomposition, the \textit{inter-spike spectrum}, naturally provides an analogue to the power spectrum for recurrence analysis in the sense that it reveals the dominant periodicities of the underlying time series. We show that the inter-spike spectrum correctly identifies patterns and transitions in the underlying system in a wide variety of examples and is robust to measurement noise.}
%
%A novel kind of power spectrum is constructed, the \textit{inter-spike spectrum}, which transforms any signal into its spike-frequency domain. This method 
%clearly shows the apparent cycles in the data and overcomes the problems for spike-train-like signals when using the obvious idea of Fourier-transforming it. 
%We invent this instructive approach with the 
%idea of transforming the $\tau$-recurrence rate of a recurrence plot (RP), which often has a spiky appearance. The $\tau$-recurrence rate is the density of recurrence points 
%along diagonals of the RP, which are parallel to the main diagonal with a distance of $\tau$. In this context the inter-spike spectrum can be interpreted as a nonlinear power 
%spectrum of a potentially high dimensional system which constitutes the RP. The proposed measure is robust to noise and is able to detect and analyze bifurcations.}

@@

The signal, which we would like to transform\footnote{We focus on $\tau$-RR in this article, Eq.~\eqref{eq_tau_rr}, since 
such time series are often spiky, which favors the proposed decomposition technique. But the proposed method can be applied to any sort of signal.} is decomposed into a set of appropriate basis functions. 
The general idea is that the sum of these weighted functions can approximate a finite signal to a sufficient degree. The weights (also called modes) corresponding to the individual basis functions can also 
be called \textit{loadings} and must be determined. A number of decomposition techniques are out there based on different sets of basis functions, e.g., 
trigonometric functions (Fourier- and wavelet analysis \cite{Bracewell1986}), 
eigenvectors of the corresponding covariance matrix (Principal Component Analysis \cite{Hotelling1933} and related techniques) 
or Intrinsic Mode Functions (Empirical Mode Decomposition and Hilbert spectrum \cite{Norden1998}).

Here we propose to use Dirac combs (DC) with 
different inter-spike periods as basis functions, Eq.~\eqref{eq_dirac_comb}. Let $s(t_i)$ be the min-max-normalized signal we want to transform of length $N$ and 
$t_i=i\cdot \Delta t,~i=1,\ldots,N$, where $\Delta t$ denotes the sampling time and $s(t_i) \in [0,\ 1]\ \forall\ i$. In the following we label this time series as a $(1\times N)$-dimensional 
vector $\vec{s}$. 
First, $\tilde{N}$ different DCs of length $N$ are constructed with inter-spike periods $T_\text{is} \in [1,\ldots,\tilde{N}]$ and $\tilde{N}=\ceil*{N/2}+1$. Second, in order to account 
for possible phase shifts of 
these basis functions occurring in $\vec{s}$, each of these $\tilde{N}$ different DCs also need to be shifted one step further $T_\text{is}-1$ times. This leaves us with a total number of 
$M = \sum_{i=1}^{\tilde{N}}i$ {\color{red}(let's come back to your example in Fig.3, where $N=4$, regarding your explanation $M=\Sigma_i^{\tilde{N}=3} i= 1+2+3=6$. So finally matrix X should be 6x4 size. However, X does not have this size in Fig.3, which is the source of confusion.)}
basis functions which we can arrange as rows of a $(M\times N)$-sized matrix $\bfX$
(Fig.~\ref{fig_tau_rr_basis_functions} illustrates the described procedure) {\color{red}(Moreover, although you claim above that number of rows in X is M in eq. (5) you say $\tilde{N}$, and in Fig. 3 it is four. You should really make it clear and cons)}:
\begin{align}
\label{eq_basis_matrix} 
\bfX_{i,j} &= \sum_{k=0}^{N}\delta\bigl(j-1-kT_\text{is}(i)-i+T_\text{is}(i) \bigr), \quad i=1,\dots,\ceil*{N/2}+1, \quad j=1,\dots,N\\
\label{eq_T_is} T_\text{is}(i)&= n, \quad \text{such that}\quad n:~ \frac{n(n-1)}{2} + 1 \leq i <  \frac{n(n+1)}{2} + 1,\quad n \in \mathbb{N}_+.
\end{align}

Note that due to the shifting of each of the basis functions of inter-spike period $T_\text{is}$, $\bfX$ is not linear independent anymore. 
Furthermore, there will be identical basis functions and also basis functions, which do not allow for an unambiguous inter-spike period, 
if we would include all $N$ possible inter-spike periods for a signal of length $N$ instead of $\ceil*{N/2}+1$ (Fig.~\ref{fig_tau_rr_basis_functions}A).
The reason is that in contrast to a trigonometric decomposition, where the Nyquist frequency marks a lower bound for the corresponding wave period, here 
the maximum considered inter-spike period is bounded by $T_{\text{is}}^{\text{max}} = \ceil*{N/2}+1$ (schematically illustrated in 
Fig.~\ref{fig_tau_rr_basis_functions}B).\\

\noindent Eventually, an under-determined linear system
\begin{equation}
\bfX^T \boldsymbol{\beta}=\vec{s}
\label{eq_linear_system}
\end{equation}  
has to be solved for $\boldsymbol\beta$, a $(M\times 1)$-sized vector carrying the loadings we are interested in. Along a variety of algorithms which can solve this problem, we are 
particularly interested in those solutions, which promote sparsity in $\boldsymbol\beta$, since our goal is to decompose the signal $\vec{s}$ into a minimal number of basis 
functions (for an excellent overview of the topic we refer to \citet{Brunton2019}). In this paper we either use the \textit{least absolute shrinkage and selection operator} 
(LASSO) \cite{Tibshirani1996} or a \textit{sequentially thresholded least squares} (STLS) regression \cite{Brunton2016,Brunton2019} to obtain a solution $\hat{\boldsymbol\beta}$. Any other 
sparse regression technique could be used. Finally, we group loadings 
which correspond to basis functions having the same period 
$T_\text{is}$ into $\hat{\boldsymbol\beta}_f$ and obtain the 
\textit{inter-spike spectrum} by simply plotting $\hat{\boldsymbol\beta}_f$ as a function of the frequency $f=T_\text{is}^{-1}$, with $T_\text{is}=\Delta t, 2\Delta t, \dots, (\ceil*{N/2}+1) \Delta t$ 
(Fig.~\ref{fig_tau_rr_dirac_comb}C, F).

@@

The proposed decomposition does not show a leakage effect by construction. However, when introducing a sparsity regularization we give up the ability to fully reconstruct the signal from the decomposition in favor of identifying the most important periodicities more clearly. This choice is mediated by a regression regularization parameter $\alpha$. Only for $\alpha = 0$ do we obtain a solution to Eq.~(\ref{eq_linear_system}) and thus a full reconstruction of the signal. Our general approach to fixing $\alpha$ is to require that we preserve most of the original signal in the sense of matching a given (Pearson) correlation coefficient $\rho$ between the original signal $\vec{s}$ and the reconstruction. The chosen regression method (in our case, LASSO or STLS) will determine how sensitive the obtained inter-spike spectrum is with respect to the regularization given by a 
selected $\rho$ (Fig.~\ref{fig_regularization}). Regarding this, in most of our applications LASSO yields more robust results, while STLS is usually faster to compute. 
We will discuss this in more detail in Section 4. Section \ref{sec_tau_rr_discussion}.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{inter-spike spectra of power grid frequency data}\label{sec_power_grid}

Power grids are large, synchronized, complex networks whose stable functioning is indispensable for modern societies.
To maintain the stability of a power grid, the balance between energy consumption and energy generation must be ensured. 
In an AC-power grid, the grid frequency is an observable variable that reflects how well this balance is satisfied. 
In this process, the grid frequency and its deviations from the nominal frequency are continuously recorded and monitored by 
the grid operators (in Europe and many parts of the world this is $50$ \si{Hz} or $60$ \si{Hz} in America and, for example, southern Japan).
For example, if there is more (less) demand than supply, the network frequency decreases (increases) compared to the nominal frequency \cite{kundur1994power}.
\\
The frequency variations can include other information, such as the functionality of control systems \cite{gorjao2020data}, the effect of fluctuations in
renewable energies (REs), demands on the grid \cite{anvari2020stochastic} and, moreover, the effect of regular dispatches due to the trading market 
\cite{meyer2020identifying}. The latter induce periodic frequency jumps. Here we look at the frequency time series for the Great Britain (GB) and Continental Europe 
(CE) (Appendix \ref{sec_power_grid_appendix} and Fig.~\ref{fig_power_grid_time_series}A,C). Clear jumps every 30 and 60 minutes are discernible and quantitatively reflected in the corresponding 
autocorrelations (Fig.~\ref{fig_power_grid_time_series}B,D). Furthermore, the autocorrelation of the CE frequency time series shows regular peaks every $15$ \si{min} 
(see Fig.~\ref{fig_power_grid_time_series}B). These peaks are caused by a mismatch of power supply and demand \cite{weissbach2009high} during dispatches. 
In most electricity grids the operation of dispatchable power plants is scheduled in 1-hour blocks, where additional (shorter) $30$ and $15$ \si{min} intervals
might exist.\\

\begin{figure}
 \centering
 \includegraphics[width=\textwidth]{./figures/fig_power_grid_spectra}
 \caption{Averaged Fourier power spectra of recorded power grid frequency time series of \textbf{A} Central Europe (CE) and \textbf{B} Great Britain (GB) (see Appendix~\ref{sec_power_grid_appendix}). 
 The corresponding averaged inter-spike spectra of the according $\tau$-RRs, Eq.~\eqref{eq_tau_rr}, are shown in \textbf{C} and \textbf{D}, respectively. Vertical red dashed lines correspond to 
 $15$, $30$ and $60$ \si{min}. For technical details on the calculation of the spectra shown, the reader is referred to the Appendix~\ref{sec_power_grid_appendix}.}  
\label{fig_power_grid_spectra}
\end{figure}

The Fourier frequency spectra for the Central Europe data in Figure~\ref{fig_power_grid_spectra}A, however, does not display sharp peaks exactly at $15$, $30$ and $60$ \si{min}, which may partly 
be explained by the leakage effect (for technical details on the calculation of 
the spectra shown, the reader is referred to the Appendix~\ref{sec_power_grid_appendix}). Especially the $30$ and $60$ \si{min} 
peaks are split into two adjacent peaks and the local minimum in between these ``double''-peaks correspond to the exact times. The Great Britain counterpart in Fig.~\ref{fig_power_grid_spectra}B 
has sharp peaks at $15$ and $30$ \si{min} and also ``local predecessor peaks'' for $30$ and $60$ \si{min} at the same positions as in panel A. 
In contrast, the $15$ \si{min} peak is completely absent from the inter-spike spectra of the $\tau$-RRs of the frequency data (Fig.~\ref{fig_power_grid_spectra}C,D), but these show sharp peaks 
at $30$ and $60$ \si{min} (there is no leakage effect due to the proposed decomposition technique). 
In front of these peaks, smaller local peaks can be seen, which correspond to the local peaks of the Fourier power spectra at $28$ and $55$ \si{min}, respectively.
Moreover, there is an additional peak at 
$40$ \si{min} for both datasets, which is absent in the Fourier spectrum and which is not a multiple of the missing $15$ \si{min} peak. The position and magnitude of the peaks in the shown 
inter-spike spectra are robust to the chosen recurrence threshold, the regression method and its regularization as well as the sampling time of the original signal.\\

We interpret the results presented as follows. It should be noted firstly that ISS is not only capable to recognize the well-known periodicity in the power grid frequency because of the dispatching time, but also the ISS demonstrate more sharp peaks at this periodicity. The missing $15$ \si{min} in the inter-spike spectra is due to the much stronger autocorrelation at $30$ and $60$ \si{min} 
(see Fig.~\ref{fig_power_grid_time_series}B,D) and because these periods are integer multiples of $15$ \si{min} the inter-spike spectra are not able to detect it and the 
sparse regression ``drags'' the $15$ \si{min} periods into the mentioned $30$ and $60$ \si{min} peaks. Moreover, the inter-spike spectra, unlike the Fourier spectra demonstrates sharp peaks exactly 
at $30$ and $60$ \si{min}, i.e. during dispatches (not valid in case of the GB dataset). Eventually, here we have found a clear sharp peak at $40$ \si{min} which can occur 
because of any regular controls in a power grid, and can be a hint to develop the existing stochastic processes to model precisely the power grid frequency \cite{gorjao2020data}.

@@

When applying the proposed inter-spike spectrum to the $\tau$-RR of a time series we expect additional frequency/period information, due to the fact that the recurrence plot (RP), 
Eq.~\eqref{eq_rp_definition}, and its corresponding $\tau$-RR, Eq.~\eqref{eq_tau_rr}, visualize the trajectory of the embedded time series in an embedding space of higher dimension. 
However, given a sufficient embedding of the time series, we would also expect that major frequencies/periods of the non-embedded time series are incorporated in the RP, its $\tau$-RR, 
and eventually in the inter-spike spectrum of the $\tau$-RR. In order to demonstrate this we apply the inter-spike spectrum to the freely available eccentricity time series of \citet{Laskar2011}. 
This astronomical computation of the orbital motion of the Earth (here we focus on the eccentricity only) has a clear expectation value of the incorporated frequencies/periods. The three leading 
eccentricity cycles of 405 kyr period, 95 kyr period and 124 kyr period are well known in palaeoclimate studies \cite{Laskar2004,Westerhold2020}. Our aim in this section is to show that the inter-spike spectrum 
of the $\tau$-RR of the embedded eccentricity time series will reflect these cycles in a similar fashion as the Fourier power spectral density of the non-embedded eccentricity time series. We will further 
show that Fourier-transforming the $\tau$-RR instead of applying the proposed inter-spike spectrum will lead to non-satisfying results, since the spiky $\tau$-RR excites a variety of harmonics in 
the corresponding Fourier spectrum (cf. Section \ref{sec_tau_rr_intro}). We use a time series which covers the past \mytilde $67$ Mio.~years (Myr), with a total length of $N=13,421$ samples and a 
sampling period of $\Delta t= 5,000$.\\
First we compute an evolutionary short time Fourier transform (FT) using a windowsize of $ws=1,000$ samplepoints ($\equiv 5$ Myr) shifted by unity and a 
Hamming window. The spectrogram reveals the expected periods mentioned, which are highlighted and clearly visible (Fig.~\ref{fig_laskar_spectra}A).

\begin{figure}
 \centering
 \includegraphics[width=\textwidth]{./figures/fig_laskar_spectra}
 \caption{\textbf{A} Evolutionary Fourier power spectra of eccentricity time series. \textbf{B} Inter-spike spectrogram of the $\tau$-recurrence rate of the eccentricity time 
 series and \textbf{C} its Fourier spectrogram. Horizontal black dashed lines highlight the analytically expected orbital periods of $405$, $124$ and $95$ kyrs. 
 For comparability, in all cases the spectra aligned in the columns of the shown plots are normalized to probabilities (sum of unity for 
 each power spectrum). For further computational details, please refer to the main text.}  
\label{fig_laskar_spectra}
\end{figure}

\noindent Then we construct the inter-spike spectrogram of the $\tau$-RR by first determining an appropriate embedding. By using a recent Tree-Embedding-Ansatz \cite{Kraemer2022} we minimize the 
\textit{false nearest neighbor} statistic \cite{hegger1999} and use the continuity statistic \cite{pecora2007} for potential delays. Eventually we obtain an $8$-dimensional embedding 
with delays $\tau s=[0, 18, 36, 48, 62, 101, 114]$ (in sampling units) for the entire time series. Similar to the preceding approach we embed the time series with these embedding parameters 
on windows of size $ws=1,200$ and a unity shift.\footnote{We use a slightly larger window here than in the short FT, because the embedding causes a ``loss'' of data points and we want to 
cover similar time spans.} The RPs are computed on these embedded trajectories with a fixed recurrence threshold corresponding to $10\%$ global recurrence rate \cite{kraemer2018} 
and the inter-spike spectra of the corresponding $\tau$-RRs is obtained by using STLS regression (see Section \ref{sec_tau_rr_method}) and a regularization threshold corresponding to 
$\rho=0.9$ accordance of $\tau$-RRs and re-composed signals. We only use the first $200$ data points of the $\tau$-RR (covering a time span of $1$ Myr). 
The spectrogram also highlights the $95$, $124$ and $405$ kyr periods as expected (Fig.~\ref{fig_laskar_spectra}B). 
Additional power is distributed in its harmonics at $190$ and $248$ kyr periods. Finally, the standard FT with a Hamming window of the same $\tau$-RRs used to obtain the inter-spike 
spectrogram in panel B, yields a smeared spectrogram which does not reflect the expected periods, but rather suffers from the spike train behavior, i.e., many excited harmonics, of the 
FT described in Section \ref{sec_tau_rr_intro} and Fig.~\ref{fig_tau_rr_dirac_comb}. The shown results are robust to a change of embedding parameters and windowsizes. However, a too low 
regularization threshold smears the clear spectrogram in panel B.

>>>>

% Abstract (Do not insert blank lines, i.e. \\)

\abstract{In recurrence analysis, the $\tau$-recurrence rate encodes the periods of the cycles of the underlying high-dimensional time series. 
It, thus, plays a similar role to the autocorrelation for scalar time-series in encoding temporal correlations. 
However, its Fourier decomposition does not have a clean interpretation. 
Thus, there is no satisfactory analogue to the power spectrum in recurrence analysis. 
We introduce a novel way to decompose the $\tau$-recurrence rate using an over-complete basis of Dirac combs together with a sparsity regularization. 
We show that this decomposition, the \textit{inter-spike spectrum}, naturally provides an analogue to the power spectrum for recurrence analysis in the sense that it reveals the dominant periodicities of the underlying time series. 
We show that the inter-spike spectrum correctly identifies patterns and transitions in the underlying system in a wide variety of examples and is robust to measurement noise.}

@@

To obtain the inter-spike spectrum, the signal, which we would like to analyse, in our case the $\tau$-RR, is decomposed into a set of appropriate basis functions. 
The general idea common to many methods is that the sum of these weighted functions can approximate a finite signal to a sufficient degree. The weights (in some contexts also called modes or loadings) corresponding to the individual basis functions must be determined.
A number of decomposition techniques based on different sets of basis functions exist, e.g., 
trigonometric functions (Fourier- and wavelet analysis \cite{Bracewell1986}), 
eigenvectors of the corresponding covariance matrix (principal component analysis \cite{Hotelling1933} and related techniques) 
or intrinsic mode functions (empirical mode decomposition and Hilbert spectrum \cite{Norden1998}).
These methods typically share the property that the basis functions form a complete basis,
the set of basis functions is linearly independent, and, thus, the weights are uniquely determined.

Here we propose to use Dirac combs (DC) with 
different inter-spike periods as basis functions, Eq.~\eqref{eq_dirac_comb}. Let $s(t_i)$ be the min-max-normalized signal we want to transform of length $N$ and 
$t_i=i\cdot \Delta t,~i=1,\ldots,N$, where $\Delta t$ denotes the sampling time and $s(t_i) \in [0,\ 1]\ \forall\ i$. In the following we label this time series as a $(1\times N)$-dimensional 
vector $\vec{s}$. 
First, $\tilde{N}$ different DCs of length $N$ are constructed with inter-spike periods $T_\text{is} \in [1,\ldots,\tilde{N}]$ and $\tilde{N}=\ceil*{N/2}+1$. Second, in order to account 
for possible phase shifts of 
these basis functions occurring in $\vec{s}$, each of these $\tilde{N}$ different DCs also need to be shifted one step further $T_\text{is}-1$ times. This leaves us with a total number of 
$M = \sum_{i=1}^{\tilde{N}}i$ 
basis functions which we can arrange as rows of a $(M\times N)$-sized matrix $\bfX$
(Fig.~\ref{fig_tau_rr_basis_functions} illustrates the described procedure):
\begin{align}
\label{eq_basis_matrix} 
\bfX_{i,j} &= \sum_{k=0}^{N}\delta\bigl(j-1-kT_\text{is}(i)-i+T_\text{is}(i) \bigr), \quad i=1,\dots,\ceil*{N/2}+1, \quad j=1,\dots,N\\
\label{eq_T_is} T_\text{is}(i)&= n, \quad \text{such that}\quad n:~ \frac{n(n-1)}{2} + 1 \leq i <  \frac{n(n+1)}{2} + 1,\quad n \in \mathbb{N}_+.
\end{align}

Note that due to the shifting of each of the basis functions of inter-spike period $T_\text{is}$, $\bfX$ is not linear independent anymore. 
Furthermore, there will be identical basis functions and also basis functions, which do not allow for an unambiguous inter-spike period, 
if we would include all $N$ possible inter-spike periods for a signal of length $N$ instead of $\ceil*{N/2}+1$ (Fig.~\ref{fig_tau_rr_basis_functions}A).
The reason is that in contrast to a trigonometric decomposition, where the Nyquist frequency marks a lower bound for the corresponding wave period, here 
the maximum considered inter-spike period is bounded by $T_{\text{is}}^{\text{max}} = \ceil*{N/2}+1$ (schematically illustrated in 
Fig.~\ref{fig_tau_rr_basis_functions}B).

Eventually, an under-determined linear system
\begin{equation}
\bfX^T \boldsymbol{\beta}=\vec{s}
\label{eq_linear_system}
\end{equation}  
has to be solved for $\boldsymbol\beta$, a $(M\times 1)$-sized vector carrying the loadings we are interested in. Along a variety of algorithms which can solve this problem, we are 
particularly interested in those solutions, which promote sparsity in $\boldsymbol\beta$, since our goal is to decompose the signal $\vec{s}$ into a minimal number of basis 
functions (for an excellent overview of the topic we refer to \citet{Brunton2019}). In this paper we either use the \textit{least absolute shrinkage and selection operator} 
(LASSO) \cite{Tibshirani1996} or a \textit{sequentially thresholded least squares} (STLS) regression \cite{Brunton2016,Brunton2019} to obtain a solution $\hat{\boldsymbol\beta}$. Any other 
sparse regression technique could be used. Finally, we group loadings 
which correspond to basis functions having the same period 
$T_\text{is}$ into $\hat{\boldsymbol\beta}_f$ and obtain the 
\textit{inter-spike spectrum} by simply plotting $\hat{\boldsymbol\beta}_f$ as a function of the frequency $f=T_\text{is}^{-1}$, with $T_\text{is}=\Delta t, 2\Delta t, \dots, (\ceil*{N/2}+1) \Delta t$ 
(Fig.~\ref{fig_tau_rr_dirac_comb}C, F).

@@

When introducing a sparsity regularization we give up the ability to fully reconstruct the signal from the decomposition in favor of identifying the most prominent spikes, and, thus, the most important periodicities more clearly. This choice is parametrized by a regression regularization parameter $\alpha$. Only for $\alpha = 0$ do we obtain a solution to Eq.~(\ref{eq_linear_system}) and, thus, a full reconstruction of the signal. Our general approach to fixing $\alpha$ is to require that we preserve most of the original signal in the sense of matching a given (Pearson) correlation coefficient $\rho$ between the original signal $\vec{s}$ and the reconstruction. The chosen regression method (in our case, LASSO or STLS) will determine how sensitive the obtained inter-spike spectrum is with respect to the regularization given by a 
selected $\rho$ (Fig.~\ref{fig_regularization}). Regarding this, in most of our applications LASSO yields more robust results, while STLS is usually faster to compute. 
We will discuss this in more detail Section \ref{sec_tau_rr_discussion}.
The proposed decomposition does not show a leakage effect by construction. However, when introducing a sparsity regularization we give up the ability to fully reconstruct the signal from the decomposition in favor of identifying the most important periodicities more clearly. This choice is mediated by a regression regularization parameter $\alpha$. Only for $\alpha = 0$ do we obtain a solution to Eq.~(\ref{eq_linear_system}) and thus a full reconstruction of the signal. Our general approach to fixing $\alpha$ is to require that we preserve most of the original signal in the sense of matching a given (Pearson) correlation coefficient $\rho$ between the original signal $\vec{s}$ and the reconstruction. The chosen regression method (in our case, LASSO or STLS) will determine how sensitive the obtained inter-spike spectrum is with respect to the regularization given by a 
selected $\rho$ (Fig.~\ref{fig_regularization}). Regarding this, in most of our applications LASSO yields more robust results, while STLS is usually faster to compute. 
We will discuss this in more detail in Section \ref{sec_tau_rr_discussion}.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Inter-spike spectra of power grid frequency data}\label{sec_power_grid}

Power grids are large, synchronized, complex networks whose stable functioning is indispensable for modern societies.
To maintain the stability of a power grid, the balance between energy consumption and energy generation must be ensured. 
In an AC-power grid, the grid frequency is an observable variable that reflects how well this balance is satisfied. 
In this process, the grid frequency and its deviations from the nominal frequency are continuously recorded and monitored by 
the grid operators (in Europe and many parts of the world this is $50$ \si{Hz} or $60$ \si{Hz} in America and, for example, southern Japan).
For example, if there is more (less) demand than supply, the network frequency decreases (increases) compared to the nominal frequency \cite{kundur1994power}.

The frequency variations can include other information, such as the functionality of control systems \cite{gorjao2020data}, the effect of fluctuations in
renewable energies (REs), demands on the grid \cite{anvari2020stochastic} and, moreover, the effect of regular dispatches due to the trading market 
\cite{meyer2020identifying}. The latter induce periodic frequency jumps. Here we look at the frequency time series for the Great Britain (GB) and Continental Europe 
(CE) (Appendix \ref{sec_power_grid_appendix} and Fig.~\ref{fig_power_grid_time_series}A,C). Clear jumps every 30 and 60 minutes are discernible and quantitatively reflected in the corresponding 
autocorrelations (Fig.~\ref{fig_power_grid_time_series}B,D). Furthermore, the autocorrelation of the CE frequency time series shows regular peaks every $15$ \si{min} 
(see Fig.~\ref{fig_power_grid_time_series}B). These peaks are caused by a mismatch of power supply and demand \cite{weissbach2009high} during dispatches. 
In most electricity grids the operation of dispatchable power plants is scheduled in 1-hour blocks, where additional (shorter) $30$ and $15$ \si{min} intervals
might exist.

\begin{figure}
 \centering
 \includegraphics[width=\textwidth]{./figures/fig_power_grid_spectra}
 \caption{Averaged evolutionary Fourier power spectra of recorded power grid frequency time series of \textbf{A} Central Europe (CE) and \textbf{B} Great Britain (GB) (see 
 Fig.~\ref{fig_power_grid_time_series} in Appendix~\ref{sec_power_grid_appendix}). 
 The corresponding averaged inter-spike spectra of the according $\tau$-RRs, Eq.~\eqref{eq_tau_rr}, are shown in \textbf{C} and \textbf{D}, respectively. Vertical red dashed lines correspond to 
 $15$, $30$ and $60$ \si{min}. For technical details on the calculation of the spectra shown, i.e preprocessing and window size being used, the reader is referred to the Appendix~\ref{sec_power_grid_appendix}.}  
\label{fig_power_grid_spectra}
\end{figure}

The evolutionary Fourier frequency spectrum for the Central Europe data in Figure~\ref{fig_power_grid_spectra}A, however, does not display sharp peaks exactly at $15$, $30$ and $60$ \si{min}, which may partly 
be explained by the leakage effect (for technical details on the calculation of 
the spectra shown, the reader is referred to the Appendix~\ref{sec_power_grid_appendix}). Especially the $30$ and $60$ \si{min} 
peaks are split into two adjacent peaks and the local minimum in between these ``double''-peaks correspond to the exact times. The Great Britain counterpart in Fig.~\ref{fig_power_grid_spectra}B 
has sharp peaks at $15$ and $30$ \si{min} and also ``local predecessor peaks'' for $30$ and $60$ \si{min} at the same positions as in panel A. 
The fact that sharp peaks are seen here using the same window size and sampling frequency as in the case of the Central European data shows that the leakage effect alone is not the cause of 
the peak splitting shown in Fig.~\ref{fig_power_grid_spectra}A.

In contrast, the $15$ \si{min} peak is completely absent from the inter-spike spectra of the $\tau$-RRs of the frequency data (Fig.~\ref{fig_power_grid_spectra}C,D), but these show sharp peaks 
at $30$ and $60$ \si{min} (there is no leakage effect due to the proposed decomposition technique). 
In front of these peaks, smaller local peaks can be seen, which correspond to the local peaks of the Fourier power spectra at $28$ and $55$ \si{min}, respectively.
Moreover, there is an additional peak at 
$40$ \si{min} for both datasets, which is absent in the Fourier spectrum and which is not a multiple of the missing $15$ \si{min} peak. The position and magnitude of the peaks in the shown 
inter-spike spectra are robust to the chosen recurrence threshold, the regression method and its regularization as well as the sampling time of the original signal.

We interpret the results presented as follows. The missing $15$ \si{min} in the inter-spike spectra is due to the much stronger autocorrelation at $30$ and $60$ \si{min} 
(see Fig.~\ref{fig_power_grid_time_series}B,D) and because these periods are integer multiples of $15$ \si{min} the inter-spike spectra are not able to detect it and the 
sparse regression ``drags'' the $15$ \si{min} periods into the mentioned $30$ and $60$ \si{min} peaks. Moreover, the inter-spike spectra, unlike the Fourier spectra demonstrates sharp peaks exactly 
at $30$ and $60$ \si{min}, i.e. during dispatches (not valid in case of the GB dataset). Eventually, here we have found a clear sharp peak at $40$ \si{min} which can occur 
because of any regular controls in a power grid, and can be a hint to develop the existing stochastic processes to model precisely the power grid frequency \cite{gorjao2020data}.

@@

When applying the proposed inter-spike spectrum to the $\tau$-RR of a time series we expect additional frequency/period information, due to the fact that the recurrence plot (RP), 
Eq.~\eqref{eq_rp_definition}, and its corresponding $\tau$-RR, Eq.~\eqref{eq_tau_rr}, visualize the trajectory of the embedded time series in an embedding space of higher dimension. 
However, given a sufficient embedding of the time series, we would also expect that major frequencies/periods of the non-embedded time series are incorporated in the RP, its $\tau$-RR, 
and eventually in the inter-spike spectrum of the $\tau$-RR. In order to demonstrate this we apply the inter-spike spectrum to the freely available eccentricity time series of \citet{Laskar2011}. 
This astronomical computation of the orbital motion of the Earth (here we focus on the eccentricity only) has a clear expectation value of the incorporated frequencies/periods. The three leading 
eccentricity cycles of 405 kyr period, 95 kyr period and 124 kyr period are well known in palaeoclimate studies \cite{Laskar2004,Westerhold2020}. Our aim in this section is to show that the inter-spike spectrum 
of the $\tau$-RR of the embedded eccentricity time series will reflect these cycles in a similar fashion as the Fourier power spectral density of the non-embedded eccentricity time series. We will further 
show that Fourier-transforming the $\tau$-RR instead of applying the proposed inter-spike spectrum will lead to non-satisfying results, since the spiky $\tau$-RR excites a variety of harmonics in 
the corresponding Fourier spectrum (cf. Section \ref{sec_tau_rr_intro}). We use a time series which covers the past \mytilde $67$ Mio.~years (Myr), with a total length of $N=13,421$ samples and a 
sampling period of $\Delta t= 5,000$.

First we compute an evolutionary short time FT using a windowsize of $ws=1,000$ samplepoints ($\equiv 5$ Myr) shifted by unity and a 
Hamming window. The spectrogram reveals the expected periods mentioned, which are highlighted and clearly visible (Fig.~\ref{fig_laskar_spectra}A).

\begin{figure}
 \centering
 \includegraphics[width=\textwidth]{figures/fig_laskar_spectra}
 \caption{\textbf{A} Evolutionary Fourier power spectra of eccentricity time series. \textbf{B} Inter-spike spectrogram of the $\tau$-recurrence rate of the eccentricity time 
 series and \textbf{C} its Fourier spectrogram. Horizontal black dashed lines highlight the analytically expected orbital periods of $405$, $124$ and $95$ kyrs. 
 For comparability, in all cases the spectra aligned in the columns of the shown plots are normalized to probabilities (sum of unity for 
 each power spectrum). For further computational details, please refer to the main text.}  
\label{fig_laskar_spectra}
\end{figure}

\noindent Then we construct the inter-spike spectrogram of the $\tau$-RR by first determining an appropriate embedding. By using a recent tree-embedding ansatz \cite{Kraemer2022} we minimize the 
\textit{false nearest neighbor} statistic \cite{hegger1999} and use the continuity statistic \cite{pecora2007} for potential delays. Eventually we obtain an $8$-dimensional embedding 
with embedding delays $\tau =[0, 18, 36, 48, 62, 101, 114]$ (in sampling units) for the entire time series. Similar to the preceding approach we embed the time series with these embedding parameters 
on windows of size $ws=1,200$ and a unity shift.\footnote{We use a slightly larger window here than in the short FT, because the embedding causes a ``loss'' of data points and we want to 
cover similar time spans.} The RPs are computed on these embedded trajectories with a fixed recurrence threshold corresponding to $10\%$ global recurrence rate \cite{kraemer2018} 
and the inter-spike spectra of the corresponding $\tau$-RRs is obtained by using STLS regression (see Section \ref{sec_tau_rr_method}) and a regularization threshold corresponding to 
$\rho=0.9$ accordance of $\tau$-RRs and re-composed signals. We only use the first $200$ data points of the $\tau$-RR (covering a time span of $1$ Myr). 
The spectrogram also highlights the $95$, $124$ and $405$ kyr periods as expected (Fig.~\ref{fig_laskar_spectra}B). 
Additional power is distributed in its harmonics at $190$ and $248$ kyr periods. Finally, the standard FT with a Hamming window of the same $\tau$-RRs used to obtain the inter-spike 
spectrogram in panel B, yields a smeared spectrogram which does not reflect the expected periods, but rather suffers from the spike train behavior, i.e., many excited harmonics, of the 
FT described in Section \ref{sec_tau_rr_intro} and Fig.~\ref{fig_tau_rr_dirac_comb}. The shown results are robust to a change of embedding parameters and windowsizes. However, a too low 
regularization threshold smears the clear spectrogram in panel B.
